Free surface computations in mantle convection models

### 2015-07-07 free-surface-paper.tex
Stuuuf

\begin{equation}
\int_\Omega q \nabla \cdot \mathbf{u} = 0
\label{eq:weak_incompressible}
\end{equation}
where $\mathbf{w}$ and $q$ are suitably chosen test functions and the integrals over 
$\Omega$ and $\partial \Omega_N$ are over the volume of the domain and the part of the boundary with stress boundary conditions, respectively.
$\mathbf{T}$ is the stress tensor, defined by
\begin{equation}
\mathbf{T} = 2 \eta \varepsilon(\mathbf{u}) - p \mathbf{I}
\end{equation}
The integral over the surface in Equation~\eqref{eq:weak_stokes} accounts for boundary stresses, 
which should be zero when evaluated on a true free surface.
We will find it analytically useful to evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
over the hydrostatic reference surface and introduce an auxilliary variable $\zeta$ which 
represents the topography of the free surface relative to that reference surface.
The equation for time evolution of the free surface is then
\begin{equation}
\frac{d \zeta}{dt} = \mathbf{u \cdot n}
\label{eq:surface_evolution}
\end{equation}
where $\mathbf{u}$ is evaluated at the free surface.
The stress on this reference surface can be approximated by the hydrostatic pressure
\begin{equation}
\mathbf{T} \approx \rho g \zeta \mathbf{I}
\label{eq:hydrostatic}
\end{equation}
where $\rho$ is the density and $g$ is gravity. Equation~\eqref{eq:weak_stokes} then becomes

@@

\section{Implementation in \texttt{Aspect} }
\subsection{Mass conservation}
\begin{itemize}
\item mass conservation requires $\int_\Omega {\bf u \cdot n} = 0$
\item not all mesh advection schemes respect this
\item for short simulations, it is not likely to make a big difference
\item for long simulations there can be significant changes in domain volume
\item finite element projections do the best job of conservation of mass
\end{itemize}
\subsection{Remeshing}
\begin{itemize}
\item ALE does not specify what happens to interior nodes
\item primary goal is preserving mesh regularity
\item we solve a laplace equation for mesh displacements
\item different boundary conditions for the laplace problem for different velocity boundary conditions
\end{itemize}
\subsection{Parallelization}
\begin{itemize}
\item mesh vertex positions are stored in a distributed vector
\item vector is redistributed with p4est on mesh refinement
\end{itemize}
\subsection{Crameri benchmarks}
more or less good go go
\section{Discussion}

>>>>

\begin{equation}
\int_\Omega q \nabla \cdot \mathbf{u} = 0
\label{eq:weak_incompressible}
\end{equation}
where $\mathbf{w}$ and $q$ are suitably chosen test functions and the integrals over 
$\Omega$ and $\partial \Omega_N$ are over the volume of the domain and the part of the boundary with stress boundary conditions, respectively.
$\mathbf{T}$ is the stress tensor, defined by
\begin{equation}
\mathbf{T} = 2 \eta \varepsilon(\mathbf{u}) - p \mathbf{I}
\end{equation}
The integral over the surface in Equation~\eqref{eq:weak_stokes} accounts for boundary stresses, 
which should be zero when evaluated on a true free surface.

Rather than analyzing finite deformation to the free surface, we will make the analytically useful
approximation of small deformations about the hydrostatic reference surface and analyze their stability.
We will therefore  evaluate the integrals in Equation~\eqref{eq:weak_stokes} 
over the hydrostatic reference surface and introduce an auxilliary variable $\zeta$ which 
represents the (small) topography of the free surface relative to that reference surface.
The equation for time evolution of the free surface is then
\begin{equation}
\frac{d \zeta}{dt} = \mathbf{u \cdot n}
\label{eq:surface_evolution}
\end{equation}
where $\mathbf{u}$ is evaluated at the free surface.
The stress on this reference surface can be approximated by the hydrostatic pressure
\begin{equation}
\mathbf{T} \approx \rho g \zeta \mathbf{I}
\label{eq:hydrostatic}
\end{equation}
where $\rho$ is the density and $g$ is gravity. Equation~\eqref{eq:weak_stokes} then becomes

@@

\section{Implementation in \texttt{Aspect} }
\subsection{Mass conservation}
\begin{itemize}
\item mass conservation requires $\int_\Omega {\bf u \cdot n} = 0$
\item not all mesh advection schemes respect this
\item for short simulations, it is not likely to make a big difference
\item for long simulations there can be significant changes in domain volume
\item finite element projections do the best job of conservation of mass
\end{itemize}
\subsection{Remeshing}
\begin{itemize}
\item ALE does not specify what happens to interior nodes
\item primary goal is preserving mesh regularity
\item we solve a laplace equation for mesh displacements
\item different boundary conditions for the laplace problem for different velocity boundary conditions
\end{itemize}
\subsection{Parallelization}
\begin{itemize}
\item mesh vertex positions are stored in a distributed vector
\item vector is redistributed with p4est on mesh refinement
\end{itemize}

\section{Results}
\begin{itemize}
\item Crameri benchmarks
\item comparison between stabilization schemes (stability)
\item comparison between stabilization schemes (accuracy)
\end{itemize}
\section{Discussion}
